The microbiota of the proximal and distal human colon

### 2017-08-18 submission/manuscript.Rmd
pat edits made

The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location. Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC). IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012, @Forbes2016, @Halfvarson2017. The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur. For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms @Benedix2010. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010. The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012, @Albenberg2014, @Donaldson2015. In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016. UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.

@@

The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014, @Harrell2012. As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

Here we aimed to address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon. Our design used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Lumenal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research @LloydPrice2016. Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated ${\theta}$YC distances from OTU abundances and compared these distances for all individuals. Again, across all patient samples we observed a range of ${\theta}$YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).

#### Stool at exit most resembles lumenal samples from the distal colon 
Next, we calculated ${\theta}$YC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller ${\theta}$YC distance between the distal lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared ${\theta}$YC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon @Eckburg2005, @deCarcer2010, @Zhang2013. Thus interpersonal variation between subjects drives the differences between samples more than sample site or location. Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances. We built the first model to classify the lumen versus mucosal samples for the proximal and distal sides, independently (Figure 4A). The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal. The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced to specific OTUs on each side.

>>>>

The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location. Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC). IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012 @Forbes2016 @Halfvarson2017. The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur. For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms @Benedix2010. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010. The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012 @Albenberg2014 @Donaldson2015. In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016. UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.

@@

The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014 @Harrell2012. As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

Here we address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon. We used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Lumenal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A) @LloydPrice2016. Samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated distances from Operational Taxonomic Unit (OTU) abundances and compared these distances for all individuals. Again, across all patient samples we observed a range of distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).

#### Fecal samples resemble lumenal samples from the distal colon 
Next, we calculated distances to examine how each sample compared to the home-collected feces. Amidst variability between patients, we did identify significantly smaller distance between the distal lumenal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the feces, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon @Eckburg2005 @deCarcer2010 @Zhang2013. Thus interpersonal variation between subjects drove the differences between samples more than sample site or location. Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of the patient's feces, and the microbes remaining on the mucosa are more distinct.

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances. We built the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure 4A). The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal. The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced to specific OTUs on each side.
